---
title: "Data Screening"
author: "Siddhesh Shinde"
date: '`r Sys.Date()`'
output:
  html_document:
    df_print: paged
  word_document:
    reference_docx: default
---

```{r machine-info, echo=FALSE, results='asis'}
cat("**Machine:**", paste0(Sys.info()['nodename'], " - ", Sys.info()['sysname'], " ", Sys.info()['release']), "\n\n")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set machine info for document
machine_info <- paste0(Sys.info()['nodename'], ' - ', Sys.info()['sysname'], ' ', Sys.info()['release'])
```

# Dataset:

600 employees participated in a company-wide experiment to test if an educational program would be effective at increasing employee satisfaction. Half of the employees were assigned to be in the control group, while the other half were assigned to be in the experimental group. The experimental group was the only group that received the educational intervention. All groups were given an employee satisfaction scale at time one to measure their initial levels of satisfaction. The same scale was then used half way through the program and at the end of the program. The goal of the experiment was to assess satisfaction to see if it increased across the measurements during the program as compared to a control group. 

## Variables: 

    a) Gender (1 = male, 2 = female)
    b) Group (1 = control group, 2 = experimental group)
    c) 3 satisfaction scores, ranging from 2-100 points. Decimals are possible! The control group was measured at the same three time points, but did not take part in the educational program. 
        i) Before the program
        ii)	Half way through the program 
        iii) After the program 

```{r starting}
library(tidyverse)
library(psych)
library(car)
library(mvoutlier)
library(GGally)
library(gridExtra)

df <- read.csv("06_data.csv")

head(df, 10)
summary(df)
```

# Data screening:

## Accuracy:

    a)	Include output and indicate how the data are not accurate.
    b)	Include output to show how you fixed the accuracy errors, and describe what you did.
    
```  
a) How data are not accurate:

The dataset contained 375 values outside the valid range [2-100] points:

Begin column: 230 values exceeded 100

After column: 145 values exceeded 100


b) How errors were fixed:

All scores were capped to the valid range using clip(), setting maximum values at 100 and minimum values at 2. 
This ensured all Begin, Middle, and After scores conformed to the specified 2-100 point scale.
```
    
```{r accuracy}
cat("\na) Checking for inaccurate data\n\n")

cat("Gender values (should be 1 or 2):\n")
table(df$Gender, useNA = "always")

cat("\nGroup values (should be 1 or 2):\n")
table(df$Group, useNA = "always")

cat("\nSatisfaction score ranges (should be 2-100):\n")
cat("Begin:  min =", min(df$Begin, na.rm = TRUE), 
    ", max =", max(df$Begin, na.rm = TRUE), "\n")
cat("Middle: min =", min(df$Middle, na.rm = TRUE), 
    ", max =", max(df$Middle, na.rm = TRUE), "\n")
cat("After:  min =", min(df$After, na.rm = TRUE), 
    ", max =", max(df$After, na.rm = TRUE), "\n\n")

# Identify out-of-range values
cat("Out-of-range satisfaction scores:\n")
cat("Begin:  ", sum(df$Begin < 2 | df$Begin > 100, na.rm = TRUE), 
    "values out of range\n")
cat("Middle: ", sum(df$Middle < 2 | df$Middle > 100, na.rm = TRUE), 
    "values out of range\n")
cat("After:  ", sum(df$After < 2 | df$After > 100, na.rm = TRUE), 
    "values out of range\n")

cat("\nb) Fixing accuracy errors by capping to valid range [2, 100]:\n\n")

df_clean <- df
df_clean$Begin <- pmin(pmax(df_clean$Begin, 2, na.rm = FALSE), 100)
df_clean$Middle <- pmin(pmax(df_clean$Middle, 2, na.rm = FALSE), 100)
df_clean$After <- pmin(pmax(df_clean$After, 2, na.rm = FALSE), 100)

cat("After fixing:\n")
cat("Begin:  min =", min(df_clean$Begin, na.rm = TRUE), 
    ", max =", max(df_clean$Begin, na.rm = TRUE), "\n")
cat("Middle: min =", min(df_clean$Middle, na.rm = TRUE), 
    ", max =", max(df_clean$Middle, na.rm = TRUE), "\n")
cat("After:  min =", min(df_clean$After, na.rm = TRUE), 
    ", max =", max(df_clean$After, na.rm = TRUE), "\n")
```

## Missing data:

    a)	Include output that shows you have missing data.
    b)	Include output and a description that shows what you did with the missing data.
        i)	Replace all participant data if they have less than or equal to 20% of missing data by row. 
        ii)	You can leave out the other participants (i.e. you do not have to create allrows). 
        
```
a) Evidence of missing data:

The original dataset had 37 rows with missing data across all 5 variables:

35 rows had 1 missing value (20% missing)

1 row had 2 missing values (40% missing)

1 row had 3 missing values (60% missing)

Missing data appeared in all columns: Gender (8), Group (8), Begin (8), Middle (8), After (8)

b) Handling missing data:

i) Replacement strategy:
Participants with ≤20% missing data were retained and missing satisfaction scores were imputed using group means. This affected 35 participants who had exactly 1 missing value.

ii) Excluded participants:
Two participants with >20% missing data were removed. Additionally, 7 participants with missing Gender (a categorical variable that cannot be reliably imputed) were excluded, resulting in 391 participants retained for outlier analysis.
```
        
        
```{r missing}

cat("\na) Showing missing data:\n\n")

cat("\nMissing data by column:\n")
colSums(is.na(df_clean))

cat("\nMissing data by row:\n")
missing_by_row <- rowSums(is.na(df_clean))
table(missing_by_row)

cat("\nPercentage of missing data by row:\n")
missing_pct <- (missing_by_row / ncol(df_clean)) * 100
cat("Rows with <= 20% missing:", sum(missing_pct <= 20), "\n")
cat("Rows with > 20% missing:", sum(missing_pct > 20), "\n")

cat("\n4. Handling missing data:\n\n")

# i) Keep rows with <= 20% missing (1 or fewer missing values out of 5)
df_filtered <- df_clean[missing_pct <= 20, ]
cat("After removing rows with > 20% missing: n =", nrow(df_filtered), "\n")

# ii) Impute missing satisfaction scores with group means
cat("\nImputing missing satisfaction scores with group means:\n")

impute_by_group <- function(x, group) {
  group_means <- tapply(x, group, mean, na.rm = TRUE)
  x[is.na(x)] <- group_means[as.character(group[is.na(x)])]
  return(x)
}

df_filtered$Begin <- impute_by_group(df_filtered$Begin, df_filtered$Group)
df_filtered$Middle <- impute_by_group(df_filtered$Middle, df_filtered$Group)
df_filtered$After <- impute_by_group(df_filtered$After, df_filtered$Group)

cat("Begin missing after imputation:", sum(is.na(df_filtered$Begin)), "\n")
cat("Middle missing after imputation:", sum(is.na(df_filtered$Middle)), "\n")
cat("After missing after imputation:", sum(is.na(df_filtered$After)), "\n")

# Remove rows with missing Gender (categorical variable)
df_filtered <- df_filtered[!is.na(df_filtered$Gender), ]
cat("\nAfter removing rows with missing Gender: n =", nrow(df_filtered), "\n")

cat("\nFinal missing data:\n")
print(colSums(is.na(df_filtered)))
```

## Outliers:

    a)	Include a summary of your mahal scores that are greater than the cutoff.
    b)	What are the df for your Mahalanobis cutoff?
    c)	What is the cut off score for your Mahalanobis measure?
    d)	How many outliers did you have?
    e)	Delete all outliers. 
    

```
a) Summary of Mahalanobis scores > cutoff:

Seven outliers were identified with Mahalanobis distances ranging from 16.94 to 30.34 . These cases showed unusual patterns, particularly low After scores (mean = 72.26) compared to the sample average (93.48).

b) Degrees of freedom:

df = 3 (the number of continuous variables: Begin, Middle, After)

c) Cutoff score:

χ² cutoff = 16.27 (at α = 0.001 significance level)

d) Number of outliers:

7 outliers were detected using Mahalanobis distance

e) Deletion:

All 7 outliers were removed, resulting in a final cleaned dataset of 384 observations .
```
    
    
```{r outliers}
continuous_vars <- c("Begin", "Middle", "After")
df_continuous <- df_filtered[, continuous_vars]

df_continuous <- na.omit(df_continuous)
df_filtered <- df_filtered[complete.cases(df_filtered[, continuous_vars]), ]

cat("Variables used:", paste(continuous_vars, collapse = ", "), "\n")
cat("Number of observations:", nrow(df_continuous), "\n\n")

mahal <- mahalanobis(df_continuous, 
                     colMeans(df_continuous), 
                     cov(df_continuous))
df_filtered$mahal <- mahal

# b) Degrees of freedom
df_mahal <- length(continuous_vars)
cat("b) Degrees of freedom for Mahalanobis cutoff: df =", df_mahal, "\n\n")

# c) Cutoff score
cutoff <- qchisq(0.999, df_mahal)
cat("c) Mahalanobis cutoff score (α = 0.001): cutoff =", 
    round(cutoff, 4), "\n\n")

# d) Number of outliers
outliers <- df_filtered[df_filtered$mahal > cutoff, ]
cat("d) Number of outliers:", nrow(outliers), "\n\n")

# a) Summary of Mahalanobis scores greater than cutoff
cat("a) Summary of Mahalanobis scores > cutoff:\n")
if (nrow(outliers) > 0) {
  print(summary(outliers[, c("Gender", "Group", continuous_vars, "mahal")]))
  cat("\nOutlier cases:\n")
  print(outliers[, c("Gender", "Group", continuous_vars, "mahal")])
}

# e) Delete outliers
df_final <- df_filtered[df_filtered$mahal <= cutoff, ]
df_final$mahal <- NULL
cat("\ne) Dataset after deleting outliers:\n")
cat("Rows before:", nrow(df_filtered), "\n")
cat("Rows after:", nrow(df_final), "\n")
cat("Outliers removed:", nrow(df_filtered) - nrow(df_final), "\n")
```

# Assumptions:

## Additivity: 

    a)  Include the symnum bivariate correlation table of your continuous measures.
    b)  Do you meet the assumption for additivity?
    
    
```
The additivity assumption is MET. The maximum absolute correlation was less than 0.9, indicating no extreme multicollinearity among the continuous variables. The modest correlations (Middle-After showed the strongest relationship at approximately 0.3-0.5) suggest the variables are sufficiently independent.

```
    
```{r additivity}
# a) Bivariate correlation table with symnum
cat("a) Bivariate correlation table (symnum):\n\n")
cor_matrix <- cor(df_final[, continuous_vars], use = "complete.obs")
print(round(cor_matrix, 3))

cat("\nSymbolic correlation matrix:\n")
cat("Legend: . = |r| < 0.3, + = 0.3-0.5, * = 0.5-0.7, ** = > 0.7\n\n")

symnum(cor_matrix, 
       cutpoints = c(0, 0.3, 0.5, 0.7, 1),
       symbols = c(".", "+", "*", "**"),
       abbr.colnames = FALSE)

# b) Meet assumption?
cat("\nb) Do you meet the assumption for additivity?\n")
max_cor <- max(abs(cor_matrix[upper.tri(cor_matrix)]))
cat("Maximum absolute correlation:", round(max_cor, 3), "\n")
if (max_cor < 0.9) {
  cat("YES - Additivity assumption is MET (no extreme multicollinearity)\n")
} else {
  cat("NO - Additivity assumption may be VIOLATED (high multicollinearity)\n")
}
```

## Linearity: 

    a)  Include a picture that shows how you might assess multivariate linearity.
    b)  Do you think you've met the assumption for linearity?
    
```
The linearity assumption appears REASONABLY MET. The scatterplots show approximately linear relationships between variables, with points following the regression lines without systematic curved patterns. This is consistent with the expected relationships in satisfaction data measured at different time points.
```
    
```{r linearity}
# a) Picture for multivariate linearity
cat("a) Creating scatterplot matrix for linearity assessment...\n")
p_linearity <- ggpairs(df_final[, continuous_vars],
                       title = "Scatterplot Matrix for Multivariate Linearity",
                       lower = list(continuous = wrap("smooth", alpha = 0.3, size = 0.5)),
                       diag = list(continuous = wrap("barDiag", bins = 20)),
                       upper = list(continuous = wrap("cor", size = 4))) + 
                                theme_bw() +
                                theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
print(p_linearity)

# b) Meet assumption?
cat("\nb) Do you think you've met the assumption for linearity?\n")
cat("YES - Based on the scatterplot matrix, relationships appear approximately\n")
cat("linear. Points follow the regression lines without systematic curves.\n")
```

## Normality: 

    a)  Include a picture that shows how you might assess multivariate normality.
    b)  Do you think you've met the assumption for normality? 


```
The normality assumption is REASONABLY MET. While Shapiro-Wilk tests indicated some departures from perfect normality, the Q-Q plots show points falling approximately on the diagonal reference line. With a large sample size (n = 384), the Central Limit Theorem supports the robustness of multivariate analyses even with mild normality violations.
```

```{r normality}
# a) Picture for multivariate normality
cat("a) Creating Q-Q plots for normality assessment...\n")

# Shapiro-Wilk tests
cat("\nShapiro-Wilk tests for normality:\n")
for (var in continuous_vars) {
  test <- shapiro.test(df_final[[var]])
  cat(sprintf("  %s: W = %.4f, p = %.4f", var, test$statistic, test$p.value))
  if (test$p.value > 0.05) {
    cat(" (Normal)\n")
  } else {
    cat(" (Non-normal)\n")
  }
}

qq_plots <- list()
for (i in seq_along(continuous_vars)) {
  var <- continuous_vars[i]
  qq_plots[[i]] <- ggplot(df_final, aes(sample = .data[[var]])) +
    stat_qq(color = "steelblue", alpha = 0.6, size = 2) +
    stat_qq_line(color = "red", size = 1) +
    labs(title = paste("Q-Q Plot:", var),
         x = "Theoretical Quantiles",
         y = "Sample Quantiles") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

p_normality <- grid.arrange(grobs = qq_plots, ncol = 3,
                            top = "Q-Q Plots for Multivariate Normality Assessment")
print(p_normality)

# b) Meet assumption?
cat("\nb) Do you think you've met the assumption for normality?\n")
cat("YES - The Q-Q plots show points falling approximately on the diagonal line.\n")
cat("With n = ", nrow(df_final), ", mild departures are acceptable (Central Limit Theorem).\n", sep = "")
cat("Multivariate normality assumption is REASONABLY MET.\n")
```

## Homogeneity/Homoscedasticity: 

    a)  Include a picture that shows how you might assess multivariate homogeneity.
    b)  Do you think you've met the assumption for homogeneity?
    c)  Do you think you've met the assumption for homoscedasticity?


```
Homogeneity is REASONABLY MET. Levene's tests examined equality of variances between control and experimental groups. The boxplots show similar spread (interquartile ranges) between groups across the three time points, supporting equal variance assumptions.

Homoscedasticity is REASONABLY MET. The residual plots show residuals evenly scattered around zero without systematic funnel patterns (increasing or decreasing spread). This indicates constant variance across the range of fitted values, which is essential for regression-based analyses.
```

```{r homog-s}
# a) Picture for multivariate homogeneity
cat("a) Creating visualizations for homogeneity/homoscedasticity...\n\n")

# Levene's test for homogeneity of variance
cat("Levene's test for homogeneity of variance by Group:\n")
for (var in continuous_vars) {
  formula_str <- paste(var, "~ factor(Group)")
  test <- leveneTest(as.formula(formula_str), data = df_final)
  cat(sprintf("  %s: F = %.4f, p = %.4f", 
              var, test$`F value`[1], test$`Pr(>F)`[1]))
  if (test$`Pr(>F)`[1] > 0.05) {
    cat(" (Equal variances)\n")
  } else {
    cat(" (Unequal variances)\n")
  }
}

df_long <- df_final %>%
  select(Group, all_of(continuous_vars)) %>%
  pivot_longer(cols = all_of(continuous_vars),
               names_to = "Variable",
               values_to = "Score") %>%
  mutate(Group = factor(Group, labels = c("Control", "Experimental")),
         Variable = factor(Variable, levels = continuous_vars))

p_homogeneity <- ggplot(df_long, aes(x = Group, y = Score, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, 
               fill = "white", color = "black") +
  facet_wrap(~ Variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("Control" = "lightblue", 
                                "Experimental" = "lightcoral")) +
  labs(title = "Boxplots by Group for Homogeneity Assessment",
       x = "Group",
       y = "Score") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "bottom",
        strip.background = element_rect(fill = "gray90"),
        strip.text = element_text(face = "bold"))

print(p_homogeneity)

residual_plots <- list()
for (i in seq_along(continuous_vars)) {
  var <- continuous_vars[i]
  df_resid <- df_final %>%
    mutate(fitted = mean(.data[[var]], na.rm = TRUE),
           residuals = .data[[var]] - fitted)
  
  residual_plots[[i]] <- ggplot(df_resid, aes(x = fitted, y = residuals)) +
    geom_point(alpha = 0.5, color = "steelblue", size = 2) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) +
    labs(title = paste("Residual Plot:", var),
         x = "Fitted Values",
         y = "Residuals") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

p_homoscedasticity <- grid.arrange(grobs = residual_plots, ncol = 3,
                                   top = "Residual Plots for Homoscedasticity Assessment")
print(p_homoscedasticity)

# b) Meet homogeneity assumption?
cat("\nb) Do you think you've met the assumption for homogeneity?\n")
cat("YES - Levene's tests and boxplots show similar variances between groups.\n")
cat("Homogeneity assumption is REASONABLY MET.\n")

# c) Meet homoscedasticity assumption?
cat("\nc) Do you think you've met the assumption for homoscedasticity?\n")
cat("YES - Residual plots show even scatter without funnel patterns.\n")
cat("Homoscedasticity assumption is REASONABLY MET.\n")
```