---
title: "Answer 02 - Siddhesh Shinde"
output:
  word_document: default
---

```{r}
# Load required packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(moments)
library(lmtest)
library(ggplot2)
library(gridExtra)
library(corrplot)
library(car)
library(seedhash)

# Set reproducible seed
gen <- SeedHashGenerator$new("Siddhesh")
seeds <- gen$generate_seeds(5)
set.seed(seeds[2])

data("mtcars")
```

```{r}
help(mtcars)
```

```{r}
head(mtcars, 10) %>%
  kable(caption = "First 10 Rows of mtcars Dataset")

cat("\nBasic Summary Statistics:\n")
summary(mtcars)
```


--- 

Develop a hypothesis

Hypothesis: "Vehicle fuel efficiency (miles per gallon) is significantly predicted by vehicle weight, number of cylinders and horsepower after controlling for other vehicle characteristics."

Based on the data and above hypothesis, we can consider the following:

Dependent Variable: 
  mpg   Miles/(US) gallon
  
Independent Variables:
  wt    Weight (1000 lbs)
  cyl   Number of cylinders
  hp    Gross horsepower

Ignorable Columns:
  disp	Displacement (cu.in.)
  drat	Rear axle ratio
  qsec	1/4 mile time
  vs	  Engine 
  am	  Transmission
	gear	Number of forward gears
  carb	Number of carburetors
  
We can ignore these above columns because our hypothesis is only concerned with the relation between a car's fuel efficiency (mpg) and its weight (wt), number of cylinder (cyl) and horsepower (hp).

---

Check for Errors & Missing Data

```{r}
# Check for missing values
missing_count <- colSums(is.na(mtcars))
missing_percent <- round(colSums(is.na(mtcars)) / nrow(mtcars) * 100, 2)

missing_summary <- data.frame(
  Variable = names(missing_count),
  Missing_Count = missing_count,
  Missing_Percent = missing_percent
)

missing_summary %>%
  kable(caption = "Missing Data Summary")

cat("\nConclusion: The mtcars dataset contains NO missing values (0 NAs).\n")
cat("Therefore, no imputation or deletion strategies are required.\n")
```


```{r}
# Check for duplicate rows
duplicates <- sum(duplicated(mtcars))
cat("Number of duplicate records:", duplicates, "\n")
cat("\nConclusion: No duplicate records found.\n")
```


```{r}
# Select relevant variables for analysis
mtcars_subset <- mtcars %>% select(mpg, cyl, hp, wt)

detect_errors <- function(var, var_name) {
  # Create a temporary data frame with the variable
  temp_df <- data.frame(value = var)
  
  # Histogram to see distribution shape
  p3 <- ggplot(temp_df, aes(x = value)) +
    geom_histogram(fill = "orange", color = "black", bins = 30) +
    labs(title = paste("Histogram of", var_name),
         x = var_name, 
         y = "Frequency") +
    theme_minimal()
  
  # Boxplot to see outliers clearly
  p4 <- ggplot(temp_df, aes(y = value)) +
    geom_boxplot(fill = "orange", alpha = 0.5) +
    labs(title = paste("Boxplot of", var_name),
         y = var_name) +
    theme_minimal()
  
  # Arrange and print the plots
  grid.arrange(p3, p4, ncol = 2)
}

# Detect error for each variable
for(var_name in names(mtcars_subset)) {
  detect_errors(mtcars_subset[[var_name]], var_name)
}
```



--- 

Clean the data

```{r}
apply(mtcars_subset, 2, function(x) { sum(is.na(x)) })
```

The mtcars dataset do not have any missing data. If it had missing data, we can handle it by either 

Checking percentage of missing data per participant:
  We can create subsets: 
    - Replaceable: Participants with ≤5% missing data. 
    - Not Replaceable: Participants with >5% missing data (often excluded or analyzed separately).
    
Multiple Imputation with mice:
  MICE (Multivariate Imputation by Chained Equations) helps guess the missing values based on relationship with other variables.
  
Univariate Outlier Detection:

```{r}
# Function to detect outliers using IQR method
detect_outliers_iqr <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
  outliers <- x < lower_bound | x > upper_bound
  return(list(outliers = outliers, 
              lower = lower_bound, 
              upper = upper_bound,
              count = sum(outliers)))
}

# Detect outliers for each variable
outlier_results <- lapply(mtcars_subset, detect_outliers_iqr)

# Create summary table
outlier_summary <- data.frame(
  Variable = names(outlier_results),
  Lower_Bound = sapply(outlier_results, function(x) round(x$lower, 2)),
  Upper_Bound = sapply(outlier_results, function(x) round(x$upper, 2)),
  Outlier_Count = sapply(outlier_results, function(x) x$count),
  Outlier_Percentage = round(sapply(outlier_results, function(x) x$count) / nrow(mtcars) * 100, 2)
)

outlier_summary %>%
  kable(caption = "Outlier Detection Summary (IQR Method: 1.5 × IQR)")

```
Multivariate Outlier Detection:

```{r}
# Select only continuous variables for outlier check
cont_vars <- mtcars_subset %>% select(mpg, hp, wt)

# Calculate distance
mahal <- mahalanobis(cont_vars,
                    colMeans(cont_vars, na.rm=TRUE),
                    cov(cont_vars, use ="pairwise.complete.obs"))
```

```{r}
cutoff <- qchisq(1 - .005, ncol(cont_vars))

cat("Degrees of Freedom:", ncol(cont_vars), "\n")
cat("Cutoff Value:", cutoff, "\n")
summary(mahal < cutoff)
```

```{r}
# Create a data frame for plotting
outlier_plot <- data.frame(
  ID = 1:length(mahal),
  Distance = mahal,
  Outlier = mahal > cutoff
)

ggplot(outlier_plot, aes(x = ID, y = Distance, color = Outlier)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_hline(yintercept = cutoff, linetype = "dashed", color = "red") +
  geom_text(aes(label = ifelse(Outlier, ID, "")), vjust = -0.5, show.legend = FALSE) +
  labs(title = "Mahalanobis Distances",
       subtitle = paste("Points above the red line (Cutoff =", round(cutoff, 2), ") are multivariate outliers"),
       y = "Distance (D^2)",
       x = "Participant ID") +
  theme_minimal() +
  scale_color_manual(values = c("black", "red"))
```

```{r}
# Eliminate outliers
mtcars_without_outliers <- subset(mtcars_subset, mahal < cutoff)

cat("\nDimensions before outlier removal:", dim(mtcars_subset)[1], "\n")
cat("Dimensions after outlier removal:", dim(mtcars_without_outliers)[1], "\n")
```

We found few outliers in the Univariate example using IQR test and no outliers based on Multivariate test using alpha of 0.05

```{r}
# Remove IQR outliers
remove_outliers_iqr <- function(data, columns) {
  
  is_outlier <- function(x) {
    Q1 <- quantile(x, 0.25, na.rm = TRUE)
    Q3 <- quantile(x, 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    lower <- Q1 - 1.5 * IQR
    upper <- Q3 + 1.5 * IQR
    return(x < lower | x > upper)
  }
  
  outlier_flags <- sapply(data[, columns], is_outlier)
  any_outlier <- apply(outlier_flags, 1, any)
  
  cat("Outlier Detection Summary:\n")
  cat("-------------------------\n")
  for(col in columns) {
    n_outliers <- sum(outlier_flags[, col])
    cat(sprintf("%-10s: %2d outliers\n", col, n_outliers))
  }
  cat("-------------------------\n")
  cat("Total rows with any outlier:", sum(any_outlier), "\n")
  cat("Rows to keep:", sum(!any_outlier), "\n\n")
  
  return(data[!any_outlier, ])
}

# Use the function
mtcars_clean <- remove_outliers_iqr(mtcars_subset, c("mpg", "cyl", "hp", "wt"))

```

---

Check Assumptions using Parametric Tests

Additivity:

```{r}
# Select only continuous variables for correlation check
cont_vars_for_cor <- mtcars_clean %>% 
  select(mpg, hp, wt)

# Calculate correlation matrix
cor_matrix <- cor(cont_vars_for_cor)

# Display correlation matrix
cor_matrix %>%
  kable(caption = "Correlation Matrix of Continuous Variables", digits = 3)

# Create a correlation plot
corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         order = "hclust",
         addCoef.col = "black",
         number.cex = 0.7,
         tl.col = "black",
         tl.srt = 45,
         title = "Correlation Matrix: Checking for Multicollinearity",
         mar = c(0,0,2,0))

# Add interpretation guide
legend("bottomright", 
       legend = c("r > 0.90: CONCERN", "r > 0.80: Monitor", "r < 0.80: OK"),
       fill = c("darkred", "orange", "lightblue"),
       cex = 0.8)
```

```{r}
# Find high correlations (excluding diagonal)
high_cor <- which(abs(cor_matrix) > 0.80 & abs(cor_matrix) < 1, arr.ind = TRUE)

if(nrow(high_cor) > 0) {
  cat("⚠ WARNING: Found", nrow(high_cor)/2, "pairs with correlations > 0.80:\n\n")
  for(i in 1:nrow(high_cor)) {
    if(high_cor[i,1] < high_cor[i,2]) {  # Avoid duplicates
      var1 <- rownames(cor_matrix)[high_cor[i,1]]
      var2 <- colnames(cor_matrix)[high_cor[i,2]]
      cor_val <- cor_matrix[high_cor[i,1], high_cor[i,2]]
      cat(sprintf("  %s <-> %s: r = %.3f\n", var1, var2, cor_val))
    }
  }
  cat("\nConsider combining these variables or selecting one from each pair.\n")
} else {
  cat("✓ GOOD: No correlations exceed 0.80.\n")
  cat("  Multicollinearity is not a concern based on correlations.\n")
}
```

Since mpg and wt have string correlation, we remove wt from out dataset.

```{r}
mtcars_additive <- mtcars_clean %>% select(mpg, cyl, hp)
```

Linearity:

```{r}
# Scatter plots for each IV vs DV
p1 <- ggplot(mtcars_additive, aes(x = cyl, y = mpg)) +
  geom_point(size = 3, color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "MPG vs. Cylinders",
       x = "Number of Cylinders",
       y = "Miles per Gallon") +
  theme_minimal()

p2 <- ggplot(mtcars_additive, aes(x = hp, y = mpg)) +
  geom_point(size = 3, color = "darkgreen", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "MPG vs. Horsepower",
       x = "Horsepower",
       y = "Miles per Gallon") +
  theme_minimal()

grid.arrange(p1, p2, nrow = 2,
             top = "Linearity Assessment: Scatter Plots with Linear Fit")
```

```{r}
cat("\n### Linearity Interpretation:\n\n")
cat("Correlation coefficients:\n")
cat("- mpg ~ cyl: r =", round(cor(mtcars_additive$mpg, mtcars_additive$cyl), 3), "\n")
cat("- mpg ~ hp:  r =", round(cor(mtcars_additive$mpg, mtcars_additive$hp), 3), "\n")

cat("All correlations are substantial (|r| > 0.7), indicating strong linear relationships.\n")
cat("Scatter plots show relatively linear patterns without obvious curvilinear trends.\n\n")
cat("Conclusion: Linearity assumption is REASONABLY MET.\n")

```

Normality:

```{r}
# Fit the regression model
model <- lm(mpg ~ cyl + hp, data = mtcars_additive)

# Extract residuals
residuals <- residuals(model)

# Shapiro-Wilk test
shapiro_test <- shapiro.test(residuals)

cat("### Shapiro-Wilk Normality Test:\n")
cat("Test Statistic (W):", round(shapiro_test$statistic, 4), "\n")
cat("p-value:", round(shapiro_test$p.value, 4), "\n\n")

if(shapiro_test$p.value > 0.05) {
  cat("Conclusion: p >0.05. We FAIL TO REJECT the null hypothesis.\n")
  cat("Residuals are approximately normally distributed. Normality assumption is MET.\n")
} else {
  cat("Conclusion: p < 0.05. We REJECT the null hypothesis.\n")
  cat("Residuals deviate from normality. Normality assumption is VIOLATED.\n")
  cat("Note: With small sample sizes (n=32), minor deviations are common.\n")
}
```

```{r}
# Q-Q plot
par(mfrow = c(2, 2))

# Q-Q plot
qqnorm(residuals, main = "Q-Q Plot: Normality Assessment")
qqline(residuals, col = "red", lwd = 2)

# Histogram of residuals
hist(residuals, breaks = 10, col = "lightblue", border = "black",
     main = "Histogram of Residuals",
     xlab = "Residuals", prob = TRUE)
lines(density(residuals), col = "red", lwd = 2)
curve(dnorm(x, mean = mean(residuals), sd = sd(residuals)), 
      add = TRUE, col = "blue", lwd = 2, lty = 2)
legend("topright", c("Actual", "Normal"), col = c("red", "blue"), 
       lty = c(1, 2), lwd = 2)

# Residuals vs Fitted
plot(fitted(model), residuals,
     main = "Residuals vs. Fitted Values",
     xlab = "Fitted Values", ylab = "Residuals",
     pch = 19, col = "darkgreen")
abline(h = 0, col = "red", lwd = 2, lty = 2)

# Box plot of residuals
boxplot(residuals, col = "lightyellow", main = "Boxplot of Residuals",
        ylab = "Residuals")
abline(h = 0, col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

Homogeneity and Homoscedasticity:

```{r}
# Extract standardized fitted values
fitvalues <- scale(model$fitted.values)
standardized <- rstudent(model)

# Create enhanced residual plot
plot_data <- data.frame(
  Fitted = as.numeric(fitvalues),
  Residuals = standardized,
  Outlier = abs(standardized) > 2
)

ggplot(plot_data, aes(x = Fitted, y = Residuals)) +
  geom_point(aes(color = Outlier), size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  geom_vline(xintercept = 0, color = "red", linewidth = 1) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "blue") +
  geom_smooth(method = "loess", se = TRUE, color = "darkgreen", linewidth = 1) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"),
                     labels = c("Within ±2", "Outside ±2")) +
  labs(title = "Residual Plot: Checking Homoscedasticity",
       subtitle = "Points should be randomly scattered with equal spread across X-axis",
       x = "Standardized Fitted Values",
       y = "Standardized Residuals",
       color = "Status") +
  annotate("text", x = min(plot_data$Fitted, na.rm=TRUE), y = 3,
           label = "✓ GOOD: Random scatter\n✗ BAD: Funnel, curve, or clusters",
           hjust = 0, vjust = 1, color = "darkgreen", size = 3.5) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
levene_cyl <- leveneTest(mtcars_additive$mpg, mtcars_additive$cyl, center = median)
print(levene_cyl)

if(levene_cyl$`Pr(>F)`[1] > 0.05) {
  cat("✓ GOOD: Levene's test p-value =", round(levene_cyl$`Pr(>F)`[1], 4), "\n")
  cat("  Variances are not significantly different across Number of Cylinders.\n")
} else {
  cat("⚠ WARNING: Levene's test p-value =", round(levene_cyl$`Pr(>F)`[1], 4), "\n")
  cat("  Variances are significantly different across Number of Cylinders.\n")
  cat("  Consider using Welch's correction or non-parametric tests.\n")
}

levene_hp <- leveneTest(mtcars_additive$mpg, mtcars_additive$hp, center = median)
print(levene_hp)

if(levene_hp$`Pr(>F)`[1] > 0.05) {
  cat("✓ GOOD: Levene's test p-value =", round(levene_hp$`Pr(>F)`[1], 4), "\n")
  cat("  Variances are not significantly different across Horsepower.\n")
} else {
  cat("⚠ WARNING: Levene's test p-value =", round(levene_hp$`Pr(>F)`[1], 4), "\n")
  cat("  Variances are significantly different across Horsepower.\n")
  cat("  Consider using Welch's correction or non-parametric tests.\n")
}
```

Levene’s Test for Homogeneity suggest that we should use Welch's correction or non-parametric tests for our Independent Variables cyl and hp.
